---
title: "Shalek CLVM"
author: "Kieran Campbell"
date: "February 2017"
output: html_document
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, fig.width = 6, fig.height = 4)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(cowplot)
library(viridis)
library(splines)
library(DT)
library(edgeR)
library(limma)
library(gplots)
library(RColorBrewer)
library(clvm)
library(goseq)



# http://stackoverflow.com/questions/18509527/first-letter-to-upper-case
firstup <- function(x) {
   substr(x, 1, 1) <- toupper(substr(x, 1, 1))
x
}

set.seed(123L)

scale_vec <- function(x) (x - mean(x)) / sd(x)

```

# Load data

Load results:

```{r load-all-data}
pcavi <- readRDS("../../data/shalek/clvm_results.rds")
sce <- readRDS("../../data/shalek/sce_shalek_clvm.rds")

x <- sce$x
xs <- sce$stimulant

# Sort the damn gene names
featureNames(sce) <- firstup(tolower(featureNames(sce)))
```



```{r debugging2, eval = F}
its <- 10:length(pcavi$elbos)

df <- data_frame(scale_vec(pcavi$elys[its]), 
                 scale_vec(pcavi$elps[its]), 
                 scale_vec(pcavi$elqs[its]), 
                 scale_vec(pcavi$elbos[its]), iter = its)#iter = 1:length(pcavi$elys))
names(df) <- c("E[p(y|T)]", "E[p(T)]", "H[q(T)]", "ELBO", "Iter")
gather(df, quantity, value, -Iter) %>%
  ggplot(aes(x = Iter, y = value, color = quantity)) +
  geom_line() +
  theme(legend.position = "top", legend.title = element_blank()) +
  labs(subtitle = "Scaled ELBO quantities") + facet_wrap(~ quantity)
```


More debugging:

```{r more-debugging, eval = F}
df <- bind_cols(pcavi[c('ts', 'cs', 'taus', 'alphas', 'betas')])
df <- tbl_df(apply(df, 2, scale_vec))
df$Iter <- 1:nrow(df)

                
gather(df, quantity, value, -Iter) %>% 
  ggplot(aes(x = Iter, y = value, color = quantity)) +
  geom_line() +
  theme(legend.position = "top", legend.title = element_blank()) +
  labs(subtitle = "Scaled ELBO quantities")
```




Plot the elbo:

```{r plot-elbo}
plot_elbo(pcavi) +
  xlab(expression("CAVI iterations"~x10^{-1})) +
  theme_cowplot(font_size = 11)

ggsave("../../figs/elbo_shalek.png", width = 5, height = 3)
```


# Exploring phenotime results

## Comparison to PC1

```{r pseudotime-vs-pc1}
pc1 <- prcomp(t(exprs(sce)), scale = TRUE)$x[,1]
 
qplot(scale_vec(pc1), pcavi$m_t, color = factor(xs)) +
  scale_color_brewer(palette = "Set1", name = "Stimulant") +
  #stat_function(fun = function(x) x, color = 'black') + 
  xlab("PC1") + ylab(expression(z))
```


## Identifying significant genes


```{r beta-df}
df_beta <- data_frame(
  c = pcavi$m_c,
  beta = pcavi$m_beta[1,],
  chi = pcavi$chi_exp[1,],
  alpha = pcavi$m_alpha[1,],
  gene = featureNames(sce),
  is_sig = significant_interactions(pcavi, n = 3)[,1]
)


df_beta$is_sig_graph <- 
  plyr::mapvalues(df_beta$is_sig, from = c(FALSE, TRUE),
                  to = c("Non-significant", "Significant"))
```

This gives:

```{r how-many-sig}
sum(df_beta$is_sig)
```

_significant_ interactions.

We can plot these:

```{r plot-sigs}
chi_cutoff <- 0.2
df_beta <- dplyr::mutate(df_beta, 
                  gene_symbol = gene)


ggplot(df_beta, aes(x = beta, y = 1 / chi, color = is_sig)) + 
  geom_point(alpha = 0.8) +
  # geom_point(data = dplyr::filter(df_beta, is_mmr), alpha = 0.6) +
  scale_color_brewer(palette = "Set1", name = "'Significant'") +
     geom_label_repel(data = dplyr::filter(df_beta, 1 / chi > chi_cutoff), aes(label = gene_symbol)) +
  xlab(expression(beta)) +
  ylab(expression(paste("[", chi, "]"^-1)))
  
```

Which genes are sig?

```{r dt-sig}
dplyr::filter(df_beta, is_sig) %>% dplyr::arrange(desc(abs(beta))) %>% 
  dplyr::select(gene, beta, alpha) %>% 
  datatable()
```

We can plot expression along phenotime for the 35 _significant_ genes:

```{r exprs-along-pst, fig.width = 12, fig.height = 9}
tmap <- pcavi$m_t


top_genes <- dplyr::filter(df_beta, is_sig) %>% 
  dplyr::arrange(desc(abs(beta))) %>% 
  extract2("gene") %>% head(n=20)

t(exprs(sce))[, top_genes] %>% 
  as_data_frame() %>% 
  dplyr::mutate(phenotime = tmap, x = factor(xs)) %>% 
  gather(gene, expression, -phenotime, -x) %>% 
  ggplot(aes(x = phenotime, y = expression, color = x)) + geom_point(alpha = 0.4) +
  facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(palette = "Set1", name = "Stimulant") +
  theme(legend.position = "top", strip.text.x = element_text(size = 9)) +
  stat_smooth(se = FALSE, method = "lm", size = 2, formula = y ~ ns(x,3))
```

Check z loosely corresponds to time:

```{r dfz}
tmap <- pcavi$m_t
zdf <- data_frame(z = tmap, time = sce$time, stimulant = xs)

ggplot(zdf, aes(x = time, y = z, fill = time)) + 
  geom_violin(alpha = 0.8) +
  theme(legend.position = "none") + 
  scale_fill_brewer(palette = "Set1") +
  # scale_fill_brewer(palette = "Set1") +
  xlab("Capture time") +
  ylab("Pathway score\n(pseudotime)") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 9),
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10)) 

time_plot <- last_plot()
```




```{r mlh1}

make_gene_plot <- function(genes) {
  inds <- match(genes, featureNames(sce))
  
  df <- t(exprs(sce))[, inds, drop=FALSE] 
  colnames(df) <- featureNames(sce)[inds]
  
  df <- as_data_frame(df) %>% 
    dplyr::mutate(phenotime = tmap, x = as.character(xs)) %>% 
    gather(gene, expression, -phenotime, -x)
  
  #df$x[is.na(df$x)] <- "NA"
  #df$x <- plyr::mapvalues(df$x, from = sort(unique(df$x)), to = c("MSI-low", "MSI-high"))
  
  #df$x <- factor(df$x, levels = c("MSI-low", "MSI-high"))
  
  df$gene <- factor(df$gene, levels = genes)
  
  df_nox <- dplyr::select(df, -x)
  
  ggplot(df, aes(x = phenotime, y = expression, fill = x, color = x)) + 
    geom_point(data = df_nox, color = "grey80", fill = "grey80") +
    geom_point(shape = 21, color = 'grey30', alpha = 0.5) +
    facet_grid(gene ~ x, scales = "free") + 
    scale_fill_brewer(palette = "Set1", name = "MSI") +
    scale_color_brewer(palette = "Set1", name = "MSI") +
    theme(legend.position = "none", strip.text = element_text(size = 10, face = "bold"),
          strip.text.y = element_text(size = 8),
          strip.background = element_rect(colour = "black", fill = "grey95", 
                                          size = 0.5, linetype = 1)) +
  theme(axis.text = element_text(size = 9),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 10)) +
    stat_smooth(se = FALSE, method = "lm", size = 1.5, formula = y ~ ns(x,3),
                color = 'grey30') +
    ylab(expression("Normalised expression")) + 
    xlab("Pathway score (pseudotime)")
}

# genes <- c("Ifi204", "Ms4a6c", "Procr")
genes <- top_genes[1:4]

make_gene_plot(genes)

gene_plot <- last_plot()
```

# Differential expression comparison

We'll use limma-voom for standard differential expression as to whether a tumour is MSI-high or not:

```{r limma}
lps_inds <- sce$stimulant == "LPS"
pam_inds <- sce$stimulant == "PAM"

wtest_results <- apply(exprs(sce), 1, function(y) {
  lps_exprs <- y[lps_inds]
  pam_exprs <- y[pam_inds]
  
  mean_lps <- mean(lps_exprs)
  mean_pam <- mean(pam_exprs)
  mean_change <- mean_lps - mean_pam
  wtest <- wilcox.test(lps_exprs, pam_exprs)
  c(mean_lps, mean_pam, mean_change, wtest$p.value)
})

wtest_results <- t(wtest_results)
wtest_results <- as_data_frame(wtest_results)
names(wtest_results) <- c("mean_expression_lps",
                          "mean_expression_pam",
                          "mean_change_lps_over_pam",
                          "p_value")
wtest_results <- dplyr::mutate(wtest_results,
                        q_value = p.adjust(p_value, method = "BH"),
                        gene = featureNames(sce))

wtest_results <- dplyr::select(wtest_results, gene, 
                               mean_expression_lps,
                              mean_expression_pam,
                        mean_change_lps_over_pam,
                        p_value, q_value)
```

And merge the results with phenotime:

```{r merge-with-us}
df_de <- inner_join(df_beta, wtest_results, by = "gene")
```


```{r compare-beta}

cols <- RColorBrewer::brewer.pal(3, "Set2")
cols2 <- c("#c5e2d9", cols[2])

ggplot(df_de, aes(x = beta, y =-log10(q_value), color = is_sig_graph)) + 
  geom_point() +
  ylab(expression(paste("Wilcoxon -", log[10], "(q-value)"))) + 
  xlab(expression(paste("PhenoPath ", beta))) +
  geom_hline(yintercept = -log10(0.05), linetype = 2, alpha = 0.5) +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 9),
        axis.title.y = element_text(size = 8)) +
  theme(legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  scale_color_manual(values = cols2, name = "Interaction") +
  theme(legend.margin = margin(), legend.position = "none")

de_plot <- ggExtra::ggMarginal(last_plot(), margins = "y", type = "histogram", size = 10)
```

Again no obvious relationship.

# Crossover bit for chris

```{r df-beta}
df_beta <- mutate(df_beta,
                  crossover = - alpha / beta)

filter(df_beta, is_sig) %>% 
  ggplot(aes(x = crossover)) + geom_histogram()
```



```{r more-vis}
textinfo <- frame_data(
  ~x, ~y, ~label,
  1, 0.2, "Gene upregulated along pseudotime\nLPS increases upregulation",
  1, -0.2, "Gene upregulated along pseudotime\nPAM increases upregulation",
  -1.1, 0.2, "Gene downregulated along pseudotime\nPAM increases downregulation",
  -1.1, -0.2, "Gene downregulated along pseudotime\nLPS increases downregulation"
)

cols <- RColorBrewer::brewer.pal(3, "Set2")
cols2 <- c("#c5e2d9", cols[2])

ggplot(df_beta, aes(x = c, y = beta, color = is_sig_graph)) + 
  geom_point() +
  geom_vline(xintercept = 0, linetype = 2, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  scale_color_manual(values = cols2, name = "Interaction") +
  geom_text_repel(data = dplyr::filter(df_beta, abs(beta) > 1.1),
                aes(label = gene), color = 'black',
                size = 2.5) +
  ylab("Covariate pathway interaction") +
  xlab("Pathway loading") +
  theme(legend.position = 'bottom',  #, legend.margin = margin(),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11),
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10)) +
  geom_text(data = textinfo, aes(x = x, y = y, label = label), 
            color = 'black', size = 3)
lplot <- ggExtra::ggMarginal(last_plot(), margins = "y", type = "histogram", size = 10)
print(lplot)
```

# What is pseudotime (2)


```{r get-correlations}
genome <- "mm9"
id <- "geneSymbol"

all_genes <- featureNames(sce)

upreg_genes <- filter(df_beta, is_sig, beta > 0) %>% extract2("gene") 
downreg_genes <- filter(df_beta, is_sig, beta < 0) %>% extract2("gene")

upg <- 1 * (all_genes %in% upreg_genes)
downg <- 1 * (all_genes %in% downreg_genes)

names(upg) <- names(downg) <- all_genes

pwfup <- nullp(upg, genome, id)
goup <- goseq(pwfup, genome, id, test.cats = "GO:BP")

pwfdown <- nullp(downg, genome, id)
godown <- goseq(pwfdown, genome, id, test.cats = "GO:BP")
```

Graph results:

```{r graph-res}
parse_go <- function(go, type, n_tested) {
  go <- go %>% 
    mutate(qval = p.adjust(over_represented_pvalue)) %>% 
    mutate(log10qval = -log10(qval),
           prop_in_cat = numInCat / n_tested) %>% 
    head(n = 12) %>% 
    mutate(type = type) %>% 
    tbl_df() %>% 
    arrange(desc(log10qval))
  go
}

gos <- bind_rows(
  parse_go(goup, "LPS up-regulated", length(upreg_genes)),
  parse_go(godown, "PAM up-regulated", length(downreg_genes))
)

gos$term <- factor(gos$term, levels = gos$term[order(gos$log10qval)])

gos <- mutate(gos, prop_cat = numDEInCat / numInCat)

filter(gos, type == "LPS up-regulated") %>% 
ggplot(aes(x = term, y = log10qval)) + #, size = 100 * prop_cat)) +
    geom_point() +
    coord_flip() +
  # facet_wrap(~ type, ncol = 1, scales = "free_y") +
    theme(axis.title.y = element_blank()) +
    ylab(expression(paste("-", log[10], " q-value"))) +
  scale_size(name = "% category\nrepresented") +
    theme(#legend.position = c(-.2, -.14), legend.direction = "horizontal",
      legend.position = c(0, 0), legend.direction = "horizontal",
        axis.text = element_text(size = 8),
        axis.title = element_text(size = 10),
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10)) 

goplot <- last_plot()
```


```{r large-plot}
lsize <- 11

bottom_top_left <- plot_grid(time_plot,NULL, de_plot, nrow = 1,
                                rel_widths = c(7,1,6),
                                # labels = c("D", "E", ""), 
                             label_size = lsize)

bottom_left <- plot_grid(bottom_top_left, goplot,
                         ncol = 1, # labels = c("", "F"),
                         label_size = lsize)

bottom_grid <- plot_grid(bottom_left, gene_plot, nrow = 1, # labels = c("", "G"),
                         label_size = lsize, rel_widths = c(4,5))

plot_grid(lplot, bottom_grid, ncol = 1, rel_heights = c(2,1.5), # labels = c("C",""),
          label_size = lsize)

ggsave("../../figs/shalek.png", width = 9, height = 8)
```

We also need one with lettering for thesis:

```{r lettering}
lsize <- 11

bottom_top_left <- plot_grid(time_plot, NULL, de_plot, nrow = 1,
                                rel_widths = c(7,1,6),
                                labels = c("", "", "C"), 
                             label_size = lsize, hjust = c(0.5, 0.5))

bottom_left <- plot_grid(bottom_top_left, goplot,
                         ncol = 1, labels = c("B", "D"),
                         label_size = lsize)

bottom_grid <- plot_grid(bottom_left, gene_plot, nrow = 1, labels = c("", "E"),
                         label_size = lsize, rel_widths = c(4,5))

pthesis <- plot_grid(lplot, bottom_grid, ncol = 1, rel_heights = c(2,1.5), labels = c("A",""),
          label_size = lsize)

ggsave("../../figs/shalek_thesis.png", pthesis, width = 9, height = 8)

```

```{r check-expression-of-genes-time-series}
d <- exprs(sce)[genes, ] %>% t() %>% 
  as_data_frame() %>% 
  mutate(time = pData(sce)[['time']], stimulant = pData(sce)[['stimulant']]) %>% 
  gather(gene, expression, -time, -stimulant)

ds <- group_by(d, time, stimulant, gene) %>% 
  summarise(mean_exprs = mean(expression))

ggplot(d, aes(x = time, y = expression, color = stimulant)) +
  geom_jitter(alpha = 0.1) + facet_wrap(~ gene) +
  scale_color_brewer(palette = "Set1") +
  geom_line(data = ds, aes(y = mean_exprs, group = stimulant))
```

Linear model:

```{r linear-model}
fit <- lm(pcavi$m_t ~ sce$time)
time_numeric <- as.numeric(gsub("h", "", sce$time))
fit2 <- lm(pcavi$m_t ~ time_numeric)
print(summary(fit))
```

DPT:

```{r dpt, eval = FALSE}
library(dpt)
ts <- Transitions(sce)
pt <- dpt(ts, branching = FALSE)
pt$tmap <- pcavi$m_t
pt$time <- sce$time

dpt_fit <- lm(DPT ~ time, data = filter(pt, DPT > 0.13))
dpt_r2 <- summary(dpt_fit)$r.squared

r2_df <- frame_data(
  ~ x, ~ y, ~ R2,
  0.132, 0.15, paste0("R^{2}~with~capture~time:~", signif(dpt_r2, 1))
)

qplot(seq_along(pt$DPT), sort(pt$DPT)) +
  xlab("Index") + ylab("DPT pseudotime (sorted)")

dpt_sort_plot <- last_plot()

filter(pt, DPT > 0.13) %>% 
  ggplot(aes(x = tmap, y =DPT, color = time)) + geom_point() +
  scale_colour_brewer(palette = "Set1", name = "Capture\ntime") +
  xlab("z") + ylab("DPT pseudotime") + 
  annotate("text", x = r2_df$x, y = r2_df$y, label = r2_df$R2, parse = TRUE) +
  theme(legend.position = 'bottom')

dpt_plot <- last_plot()
```

Monocle:

```{r monocle, eval = FALSE}
library(monocle)
sce@logged <- TRUE
cds <- toCellDataSet(sce)
cds <- reduceDimension(cds)
cds <- orderCells(cds)

pt <- pt %>% mutate(monocle = cds$Pseudotime)

monocle_fit <- lm(monocle ~ time, data = pt)
monocle_r2 <- summary(monocle_fit)$r.squared
```

```{r plot-monocle, eval = FALSE}
r2_df <- frame_data(
  ~ x, ~ y, ~ R2,
  0.16, 30, paste0("R^{2}~with~capture~time:~", signif(monocle_r2, 1))
)

ggplot(pt, aes(x = tmap, y = monocle, color = time)) + geom_point() +
  scale_colour_brewer(palette = "Set1", name = "Capture\ntime") +
  xlab("z") + ylab("Monocle pseudotime") +
  annotate("text", x = r2_df$x, y = r2_df$y, label = r2_df$R2, parse = TRUE) +
  theme(legend.position = 'bottom')

monocle_plot <- last_plot()
```

```{r put_altogether, eval = FALSE}
plot_grid(dpt_sort_plot, dpt_plot, monocle_plot, 
          labels = c("A", "B", "C"), nrow = 1)
ggsave("../../figs/s_compare_monocle_dpt.png", width = 14, height = 4)
```

# Technical

```{r technical}
sessionInfo()
```
