---
title: "PhenoPath simulation study"
author: "Kieran Campbell"
date: "March 2017"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, fig.width = 7, fig.height = 4)

library(scater)
library(limma)
library(edgeR)
library(clvm)
library(ggbeeswarm)
library(cowplot)
library(viridis)
library(readr)
library(dplyr)
library(splines)


set.seed(123L)

scale_vec <- function(x) (x - mean(x)) / sd(x)

select <- dplyr::select
filter <- dplyr::filter
mutate <- dplyr::mutate
arrange <- dplyr::arrange
```

# Load data and convert into SCESet


```{r load-all-data}
kallisto_dirs <- dir("../../data/simulations/quant", full.names = TRUE)

samples <- sapply(strsplit(kallisto_dirs, "_"), `[`, 2)

sce <- readKallistoResults(directories = kallisto_dirs, samples = samples)

fdata <- read_csv("../../data/simulations/gene_pars.csv")
pdata <- read_csv("../../data/simulations/pdata.csv")

sce$sample_index <- as.numeric(sampleNames(sce))
mm <- match(sce$sample_index, pdata$sample)
pdata <- pdata[mm, ]  
pData(sce) <- cbind(pData(sce), pdata)

sce <- plotPCA(sce, colour_by = "x", ncomponents = 3, return_SCESet = TRUE)
plotPCA(sce, colour_by = "pst", ncomponents = 3)

fData(sce) <- cbind(fData(sce), fdata)
```

# Model inference

## Limma voom

```{r limma-voom}
dge <- DGEList(counts(sce))
dge <- calcNormFactors(dge)

design <- model.matrix(~ (x == 1), pData(sce))
v <- voom(dge, design, plot = TRUE)

fit <- lmFit(v, design)
fit <- eBayes(fit)
results <- decideTests(fit)

qvals <- p.adjust(fit$p.value[,2], method = 'BH')

df_limma <- data_frame(limma_coef = fit$coefficients[,2], 
                       limma_qval = qvals,
                       feature = featureNames(sce))
df_limma <- mutate(df_limma, limma_signif = limma_qval < 0.05)
```

## Phenopath

```{r phenopath}
y <- scale(t(exprs(sce)))
x <- cbind(pData(sce)[[ 'x' ]])


pcavi <- clvm(y, x, elbo_tol = 1e-10)

pc1 <- prcomp(t(exprs(sce)))$x[,1]

pp_df <- data_frame(pp_alpha = pcavi$m_alpha[1,],
                    pp_beta = pcavi$m_beta[1,],
                    pp_c = pcavi$m_c,
                    pp_signif = as.vector(significant_interactions(pcavi, n = 2)),
                    feature = featureNames(sce),
                    alpha = fData(sce)[['alpha']],
                    beta = fData(sce)[['beta']],
                    c = fData(sce)[['c']],
                    regime = fData(sce)[['regime']])
```


## Merge results

```{r merge-results}
pp_df <- inner_join(pp_df, df_limma, by = 'feature')

regime_texts <- c("Differential\nexpression only",
                  "Pseudotime\nregulation only",
                  "Pseudotime and\ncovariate interactions",
                  "Differential\nexpression, pseudotime\nand covariate interactions")

pp_df$regime_txt <- plyr::mapvalues(pp_df$regime, 
                                    from = c("de", "pst", "pst_beta", "de_pst_beta"),
                                    to = regime_texts)

pp_df$regime_txt <- factor(pp_df$regime_txt, levels = regime_texts)

pp_pst <- data_frame(phenopath = pcavi[['m_t']],
                     pc1, pst = pData(sce)[['pst']])
```

# Results

## Effect sizes under different conditions

```{r limma-plot}
ggplot(pp_df, aes(x = regime_txt, y = abs(limma_coef), color = limma_signif)) + 
  geom_quasirandom(dodge.width = 0.5, alpha = 0.8) +
  xlab("Simulation regime") +
  ylab("Absolute value of limma coefficient") +
  scale_color_brewer(palette = "Set2",
                     name = "Limma voom\nsignificant at 5% FDR") +
  theme_cowplot(font_size = 11) +
  theme(legend.position = "top") +
  theme(axis.text.x = element_text(size = 9),
        axis.title.y = element_text(size = 9))
limma_plot <- last_plot()
```

```{r phenopath-plot}
ggplot(pp_df, aes(x = regime_txt, y = abs(pp_beta), color = pp_signif)) + 
  geom_quasirandom(dodge.width = 0.5, alpha = 0.8) +
  xlab("Simulation regime") +
  ylab(expression(paste("|", beta, "|"))) +
  scale_color_brewer(palette = "Set2",
                     name = expression(paste("PhenoPath\nsignificant at 2", sigma))) +
  theme_cowplot(font_size = 11) +
  theme(legend.position = "top") +
  theme(axis.text.x = element_text(size = 9))
phenopath_plot <- last_plot()
```

## Tabular results

```{r tab-results}
pp_table <- group_by(pp_df, regime_txt) %>% 
  summarise(n = n(), n_sig = sum(pp_signif), prop_sig = mean(pp_signif))

lm_table <- group_by(pp_df, regime_txt) %>% 
  summarise(n = n(), n_sig = sum(limma_signif), prop_sig = mean(limma_signif))
```

PhenoPath results:

```{r phenopath-results}
DT::datatable(pp_table)
```

```{r limma-results}
DT::datatable(lm_table)
```

"Case four" overlap table for supplementary:

```{r case-four}
pp_df_4_only <- filter(pp_df, regime_txt == "Differential\nexpression, pseudotime\nand covariate interactions")

overlap_df <- frame_data(
  ~regime, ~n,
  "both", sum(pp_df_4_only$pp_signif & pp_df_4_only$limma_signif),
  "PhenoPath only", sum(pp_df_4_only$pp_signif & !pp_df_4_only$limma_signif),
  "Limma only", sum(!pp_df_4_only$pp_signif & pp_df_4_only$limma_signif),
  "neither", sum(!pp_df_4_only$pp_signif & !pp_df_4_only$limma_signif)
)
```

Print results:

```{r print-overlap-results}
DT::datatable(overlap_df)
```

And lazy latex:

```{r lazy-latex, results = 'asis'}
print(xtable::xtable(overlap_df))
```

## Inferred pseudotimes

```{r pc1-plot}
ggplot(pp_pst, aes(x = pst, y = pc1, color = factor(x))) + 
  geom_point(alpha = 0.8) +
  scale_colour_brewer(palette = "Set1", name = "x") +
  theme_cowplot(font_size = 11) +
  xlab("True pseudotime") + ylab("PC1")
pc1_plot <- last_plot()
```

```{r z-plot}
ggplot(pp_pst, aes(x = pst, y = phenopath, color = factor(x))) + 
  geom_point(alpha = 0.8) +
  scale_colour_brewer(palette = "Set1", name = "x") +
  theme_cowplot(font_size = 11) +
  xlab("True pseudotime") + ylab("PhenoPath z") +
  theme(legend.position = "none")

z_plot <- last_plot()
```

## PCA representation

```{r pca-representation}
pca_df <- as_data_frame(redDim(sce)[,1:2]) %>% 
  mutate(x = factor(pData(sce)[['x']]), pst = pp_pst[['pst']])

ggplot(pca_df, aes(x = PC1, y = PC2, color = pst, shape = x)) +
  geom_point(size = 2.5) + scale_color_viridis(name = "True\npseudotime") +
  theme_cowplot(font_size = 11)

pca_plot <- last_plot()
```

We can also quantify the correlation to true pseudotime for both samples:

```{r quantify-correlation}
correlation_df <- data_frame(
  pc1 = cor(pp_pst$pst, pp_pst$pc1),
  z = cor(pp_pst$pst, pp_pst$phenopath)
)

knitr::kable(correlation_df)
```

## False discovery rate analysis

We can find approximate false positive and negative rates for both Limma Voom and PhenoPath given the effect sizes stated. We're interested in

$$ FDR = \frac{\text{# false discoveries}}{\text{# all discoveries}} $$

We can construct a helper function for learning FDRs, etc:

```{r fdr}
calculate_statistics <- function(tbl) {
  P <- sum(tbl[,2])
  N <- sum(tbl[,1])
  TP <- tbl[2,2]
  TN <- tbl[1,1]
  FP <- tbl[1,2]
  FN <- tbl[2,1]
  TPR <- TP / P
  FPR <- FP / N
  FDR <- FP / (TP + FP)
  data.frame(P = P, N = N, TP = TP, TN = TN, FP = FP, 
             FN = FN, TPR = TPR, FPR = FPR, FDR = FDR)
}
```

### Limma Voom

```{r fdr-limma}
tbl_limma <- table(pp_df$alpha != 0, pp_df$limma_signif)
limma_stats <- calculate_statistics(tbl_limma)
knitr::kable(limma_stats)
```

### PhenoPath

```{r fdr-phenopath}
tbl_phenopath <- table(pp_df$beta != 0, pp_df$pp_signif)
phenopath_stats <- calculate_statistics(tbl_phenopath)
knitr::kable(phenopath_stats)
```

# Save figures for paper

## Figure for supplementary

```{r supplementary-figure}
comp_plot <- plot_grid(pc1_plot, z_plot, rel_widths = c(6,5))

comp_plot_2 <- plot_grid(NULL, comp_plot, NULL, rel_heights = c(1,5,1),
                         labels = c("B", "", ""), ncol = 1)

left_grid <- plot_grid(pca_plot, comp_plot_2, ncol = 1, labels = c("A", ""))
right_grid <- plot_grid(limma_plot, phenopath_plot, ncol = 1,
                        labels = c("C", "D"))

plot_grid(left_grid, right_grid, nrow = 1, rel_widths = c(4,5))
ggsave("../../figs/simulations_all.png", width = 12, height = 8)
```

## Example gene figure

```{r example-gene-figure}
fData(sce)$regime <- pp_df$regime
sce$x <- factor(sce$x)

sce2 <- sce
sce2 <- sce2[, order(as.numeric(sampleNames(sce2)))]
count_mat <- read_csv("../../data/simulations/count_matrix.csv")
count_mat <- as.matrix(count_mat)
colnames(count_mat) <- sampleNames(sce2)
rownames(count_mat) <- featureNames(sce2)
exprs(sce2) <- log2(count_mat)

genes <- c(1, 101, 201, 301) + 2
sce_marker <- sce2[genes,]
regimes <- unique(pp_df$regime_txt)
regime_labels <- paste0("(", 1:4, ") ", regimes)

featureNames(sce_marker) <- regime_labels

zplt <- plotExpression(sce_marker, features  = 1:4, 
               x = "pst", colour_by = "x",
               show_violin = FALSE, exprs_values = "exprs",
               scales = "free_y", ncol = 4) + 
  # stat_smooth(se = F, method = "lm", formula = y ~ splines::ns(x, 3)) +
  xlab("Pathway score, z") +
 theme(strip.background = element_blank(),
       strip.text.x = element_blank(),
       legend.position = "none") +
  ylab(expression(paste("Expression (", log[2], "(count))")))

xplt <- plotExpression(sce_marker, features  = 1:4, 
               x = "x", colour_by = "x",
               show_violin = TRUE, exprs_values = "exprs",
               scales = "free_y", ncol = 4)  +
  theme(legend.position = "none") + xlab("Covariate, x") +
  ylab(expression(paste("Expression (", log[2], "(count))")))

plot_grid(xplt, zplt, nrow = 2, rel_heights = c(3,2))

ggsave("../../figs/simulation_example_genes.png", width = 8, height = 5)
```

## Subfigure of main text

```{r simulation-all-plot}
plot_grid(limma_plot, phenopath_plot, nrow = 2, align = 'v')

ggsave("../../figs/simulations.png", width = 9, height = 6, scale = 0.92)
```