---
title: "Immune-MSI interactions in colorectal cancer using Variational Bayes Phenotime"
author: "Kieran Campbell"
date: "February 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.width = 6, fig.height = 4)

library(scater)
library(tidyverse)
library(magrittr)
library(matrixStats)
library(ggrepel)
library(cowplot)
library(viridis)
library(splines)
library(DT)
library(edgeR)
library(limma)
library(gplots)
library(RColorBrewer)
library(clvm)

set.seed(123L)

scale_vec <- function(x) (x - mean(x)) / sd(x)

```

# Load data

Load results:

```{r load-all-data}
pcavi <- readRDS("../../data/COAD/clvm_results.rds")
sce <- readRDS("../../data/COAD/sce_coad_clvm.rds")

x <- sce$x
```

# Exploring phenotime results

## Comparison to PC1

```{r pseudotime-vs-pc1}
pc1 <- prcomp(t(exprs(sce)), scale = TRUE)$x[,1]

qplot(scale_vec(pc1), scale_vec(pcavi$m_t), color = factor(x)) +
  scale_color_brewer(palette = "Set1", name = "MSI-high") +
  stat_function(fun = function(x) x, color = 'black') + 
  xlab("PC1") + ylab(expression(z))
```

## Identifying significant genes

In the context of CAVI we have modelled each interaction coefficient by an approximating distribution of the form

$$ \beta_{pg} \sim \text{Normal}(m_{\beta_{pg}}, s_{\beta_{pg}}^2) $$
We define a _significant_ interaction if the $2 \sigma$ posterior doesn't cross zero, ie

$$ m_{\beta_{pg}} - n s_{\beta_{pg}} > 0 $$
or
$$ m_{\beta_{pg}} + n s_{\beta_{pg}} < 0 $$

Here we use $n=3$:

```{r beta-df}
df_beta <- data_frame(
  c = pcavi$m_c,
  beta_msi = pcavi$m_beta[1,],
  chi_msi = pcavi$chi_exp[1,],
  alpha_msi = pcavi$m_alpha[1,],
  gene = featureNames(sce),
  is_mmr = fData(sce)$is_mmr,
  is_sig = significant_interactions(pcavi, n = 3)
)
```

This gives:

```{r how-many-sig}
sum(df_beta$is_sig)
```

_significant_ interactions.

We can plot these:

```{r plot-sigs}
chi_cutoff <- 1
df_beta <- mutate(df_beta, 
                  gene_symbol = sapply(strsplit(gene, "_"), `[`, 1))

is_na <- df_beta$gene_symbol == "NA"
df_beta$gene_symbol[is_na] <- df_beta$gene[is_na]

ggplot(df_beta, aes(x = beta_msi, y = 1 / chi_msi, color = is_sig)) + 
  geom_point(alpha = 0.8) +
  # geom_point(data = dplyr::filter(df_beta, is_mmr), alpha = 0.6) +
  scale_color_brewer(palette = "Set1", name = "'Significant'") +
    geom_label_repel(data = dplyr::filter(df_beta, 1 / chi_msi > chi_cutoff), aes(label = gene_symbol)) +
  xlab(expression(beta)) +
  ylab(expression(paste("[", chi, "]"^-1)))
  
```

Which genes are sig?

```{r dt-sig}
filter(df_beta, is_sig) %>% arrange(desc(abs(beta_msi))) %>% 
  select(gene, is_mmr, beta_msi, alpha_msi) %>% 
  datatable()
```

We can plot expression along phenotime for the 35 _significant_ genes:

```{r exprs-along-pst, fig.width = 12, fig.height = 9}
tmap <- pcavi$m_t


top_genes <- dplyr::filter(df_beta, is_sig) %>% 
  arrange(desc(abs(beta_msi))) %>% 
  extract2("gene") %>% head(n=20)

t(exprs(sce))[, top_genes] %>% 
  as_data_frame() %>% 
  dplyr::mutate(phenotime = tmap, x = factor(x)) %>% 
  gather(gene, expression, -phenotime, -x) %>% 
  ggplot(aes(x = phenotime, y = expression, color = x)) + geom_point(alpha = 0.4) +
  facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(palette = "Set1", name = "MSI") +
  theme(legend.position = "top", strip.text.x = element_text(size = 9)) +
  stat_smooth(se = FALSE, method = "lm", size = 2, formula = y ~ ns(x,3))
```


What is phenotime? Check the immune response genes:

```{r immune-resp, fig.width = 7}
immune_genes <- c("FOXP3", "CD4", "IL2RB")
immune_inds <- match(immune_genes, fData(sce)$hgnc_symbol)

t(exprs(sce))[, immune_inds, drop=FALSE] %>% 
  as_data_frame() %>% 
  dplyr::mutate(phenotime = tmap, x = factor(x)) %>% 
  gather(gene, expression, -phenotime, -x) %>% 
  ggplot(aes(x = phenotime, y = expression, color = x)) + geom_point(alpha = 0.4) +
  facet_wrap(~ gene, scales = "free_y") + 
  scale_color_brewer(palette = "Set1", name = "MSI") +
  theme(legend.position = "top", strip.text.x = element_text(size = 9)) +
  stat_smooth(se = FALSE, method = "lm", size = 2)#, formula = y ~ ns(x,3))
```

Nice immune response plot:

```{r nice-immune-response}
df_immune <- t(exprs(sce))[, immune_inds, drop=FALSE] %>% 
  as_data_frame() %>% 
  dplyr::mutate(phenotime = tmap, x = factor(x)) %>% 
  gather(gene, expression, -phenotime, -x)

df_immune <- mutate(df_immune, 
                  gene_symbol = sapply(strsplit(gene, "_"), `[`, 1))

ggplot(df_immune, aes(x = phenotime, y = expression, color = gene_symbol)) + 
  geom_point(alpha = 0.4) +
  facet_wrap(~ gene_symbol, scales = "free_y") + 
  scale_color_brewer(palette = "Set1", name = "Gene") +
  theme(legend.position = "none", strip.text.x = element_text(size = 9)) +
  stat_smooth(se = FALSE, method = "lm", size = 2, formula = y ~ ns(x,3)) +
  ylab(expression(paste("Expression log"[2], "(TPM+1)"))) +
  xlab(expression(z))

```




Look at MLH1 in more detail:

```{r mlh1}
mlh1 <- c("MLH1")

mlh1_ind <- match(mlh1, fData(sce)$hgnc_symbol)

df_mlh1 <- t(exprs(sce))[, mlh1_ind, drop=FALSE] %>% 
  as_data_frame() %>% 
  dplyr::mutate(phenotime = tmap, x = as.character(x)) %>% 
  gather(gene, expression, -phenotime, -x)

df_mlh1$x[is.na(df_mlh1$x)] <- "NA"
df_mlh1$x <- plyr::mapvalues(df_mlh1$x, from = sort(unique(df_mlh1$x)), to = c("MSI-low", "MSI-high"))

df_mlh1$x <- factor(df_mlh1$x, levels = c("MSI-low", "MSI-high"))

df_mlh1_nox <- dplyr::select(df_mlh1, -x)

ggplot(df_mlh1, aes(x = phenotime, y = expression, color = x)) + 
  geom_point(data = df_mlh1_nox, color = "grey70") +
  geom_point() +
  facet_wrap(~ x, scales = "free_y") + 
  scale_color_brewer(palette = "Set1", name = "MSI") +
  theme(legend.position = "none", strip.text.x = element_text(size = 9)) +
  stat_smooth(se = FALSE, method = "lm", size = 1.5, formula = y ~ ns(x,3)) +
  ggtitle("MLH1 expression") +
  ylab(expression(paste("Expression log"[2], "(TPM+1)"))) +
  xlab("z (immune response)")

```

# Differential expression comparison

We'll use limma-voom for standard differential expression as to whether a tumour is MSI-high or not:

```{r limma}
# retained_fnames <- rownames(sc_tumour[rowSums(counts(sc_tumour)) > 20, ])
dge <- DGEList(counts(sce))
dge <- calcNormFactors(dge)

sce$msi_cov <- 1 * (sce$msi_status == "msi-h")
design <- model.matrix(~ msi_cov, pData(sce))
v <- voom(dge, design, plot = TRUE)

fit <- lmFit(v, design)
fit <- eBayes(fit)
results <- decideTests(fit)
# vennDiagram(results)
```

And merge the results with phenotime:

```{r merge-with-us}
int_df_limma <- dplyr::rename(df_beta, feature = gene)

df_limma <- data_frame(msi_coef = fit$coefficients[,2], 
                       msi_pval = fit$p.value[,2],
                       mu = pcavi$m_mu,
                       feature = featureNames(sce)) %>% 
  left_join(int_df_limma, by = "feature") %>% 
  dplyr::mutate(msi_log10pval = -log10(msi_pval)) 
```

Then we can compare in several ways. Firstly, are the $\alpha$ values consistent with the limma voom reported coefficients?

```{r compare-limma-alpha}
ggplot(df_limma, aes(x = msi_coef, y = alpha_msi)) + 
  geom_point(alpha = 0.5) +
  xlab("Limma voom coefficient") + ylab(expression(paste("Phenotime ", alpha)))

ggplot(df_limma, aes(x = msi_coef, y = alpha_msi, color = beta_msi)) + 
  geom_point(alpha = 0.5) +
  xlab("Limma voom coefficient") + 
  ylab(expression(paste("Phenotime ", alpha))) +
  scale_color_viridis()
```

There is some "twisting" there which may be due to using a mixed effects model rather than straight forward DE.

Is there any relationship between the limma voom reported coefficients and phenotimes $\beta$ values?

```{r compare-beta}
ggplot(df_limma, aes(x = msi_coef, y = beta_msi)) + geom_point(alpha = 0.5) +
  xlab("Limma voom coefficient") + ylab(expression(paste("Phenotime ", beta)))
```

So no obvious correlations. Let's just double check the $p$-values:

```{r limma-pval}
ggplot(df_limma, aes(x = beta_msi, y = msi_log10pval)) + geom_point(alpha = 0.5) +
  ylab("Limma voom -log10 pval") + xlab(expression(paste("Phenotime ", beta)))
```

Again no obvious relationship.

# Crossover bit for chris

```{r df-beta}
df_beta <- mutate(df_beta,
                  crossover = - alpha_msi / beta_msi)

filter(df_beta, is_sig) %>% 
  ggplot(aes(x = crossover)) + geom_histogram()
```

